---
title: "Fertility of Arabidopsis thaliana"
author: "Cyril Matthey-Doret"
date: "26 novembre 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(knitr);library(pander)
library(MASS);library(ggplot2);library(lme4);library(car)
ara <- read.csv(file = "../arabidopsis.csv")
ara$gen <- as.factor(ara$gen)
ara$rack <- as.factor(ara$rack)
ara$nutrient <- as.factor(ara$nutrient)

```

#Introduction

Here, I analyse a dataset on Arabidopsis fertility containing different factors. The aim is to measure the effect of those factors on the fertility.

#Data

The variables studied are:

  * __reg__ (factor): Region where the plant has been collected. There are 3 regions (`NL`: Netherlands, `SP`: Spain, `SW`: Sweden).
  * __popu__ (factor): Populations of origin of the plant inside each region. There are 9 levels.
  * __gen__ (factor): The plant's genotype. There are 24 different genotypes in total.
  * __rack__ (factor): The rack the plant was stored on, inside the greenhouse. There are 2 racks.
  * __nutrient__ (factor): Nutrition treatment. 2 levels (`1`: minimal nutrients, `8`: additional nutrients)
  * __amd__ (factor): Status of simulated herbivory. 2 levels (`clipped`: positive, `unclipped`: negative)
  * __status__ (factor): Germination method. 3 levels (`Normal`, `Petri.Plate`, `Transplant`)
  * __total.fruits__ (integer): Number of fuits per plant. This is the response variable, we use it to measure fertility.
  
All predictors are factors. Among predictors, `status` and `amd` are considered to be interesting, while `rack` and `status` are described as "nuisance factors".

The dataset contains 625 individual plants. 

The variable of interest being the number of fruit, the data is not normally distributed __(Figure 1)__. Indeed, a quick look at the data indicate that the values are discrete and contain `r length(ara$total.fruits[ara$total.fruits==0])` zeros, out of the `r length(ara$total.fruits)` observations. 

Some factors (`rack`, `nutrient` and `gen`) were encoded in numbers. Their type needs to be set to `Factor` in order to prevent functions from treating them as integer variables in downstream analyses. Groups are also unbalanced for certain factors. For example, there are 11 plants of genotype 5 and 47 of genotype 35.

From the repartition of levels, it appears that `popu` is nested within `reg`, as each population is confined to a single region __(Table 1)__ and `gen` is nested within `popu` __(Table 2)__.

#Analyses

As the response is not normally distributed and the design is unbalanced, it is not optimal to use regular methods such as ANOVAs.
I chose to use a generalized linear mixed model, fitting the data to a Poisson distribution, as the response variable is discrete and all values are positive or zero. The data does not fit Poisson very well __(Figure 2)__, but it is still the logical choice for a count variables. The model is as follows:

```{r model1, echo=T}
mdl1 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|status)+(1|reg/popu/gen)+(1|rack),
              family= poisson)
```

As they were stated to be of particular interes, nutrient and amd have been set to fixed factors, while the other "nuisance" variables have been set as random. This choice is rather arbitrary and could have been different if we were interested, for example in the status predictor.

Perhaps a more simple model could be more appropriate, as complex model tend to generate noise. This can be tested by iterative process, removing random factors from the model and looking at the effect of the AIC value, a relative measure of loss of information. The lower AIC, the better the model.

```{r testmod, echo=T}
mdl2 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|status)+(1|reg/popu/gen),
              family= poisson)  # rack removed
mdl3 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|reg/popu/gen)+(1|rack),
              family= poisson)  # status removed
mdl4 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|status)+(1|reg/popu)+(1|rack),
              family= poisson)  # gen removed
mdl5 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|status)+(1|popu/gen)+(1|rack),
              family= poisson)  # reg removed
mdl6 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|status)+(1|reg/gen)+(1|rack),
              family= poisson)  # popu removed
```


Among all these models, `mdl5` has the lowest AIC (17523.93), indicating the model is improved when removing the region factor.

```{r AIC_comp}

AIC.table <- AIC(mdl1,mdl2,mdl3,mdl4,mdl5,mdl6)
AIC.table$dropped <- c("None","rack","status","gen","reg","popu")
AIC.table <- AIC.table[,c(3,1,2)]
print(AIC.table)

```

We can try removing another factor from model 5: 

```{r testmod2, echo=T}
mdl5.1 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|popu/gen)+(1|rack),
              family= poisson)  # reg  and status removed
mdl5.2 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|status)+(1|gen)+(1|rack),
              family= poisson)  # reg  and popu removed
mdl5.3 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|status)+(1|popu)+(1|rack),
              family= poisson)  # reg and gen removed
mdl5.4 <- glmer(data = ara,formula = total.fruits ~ nutrient*amd+(1|status)+(1|popu/gen),
              family= poisson)  # reg and rack removed


```

```{r AIC_comp2}

AIC.table <- AIC(mdl5.1,mdl5.2,mdl5.3,mdl5.4)
AIC.table$dropped <- c("status","popu","gen","rack")
AIC.table <- AIC.table[,c(3,1,2)]
print(AIC.table)

```

Removing any additional factor increases the AIC, therefore we cannot improve the model by further removing random factors.

#Results:

```{r output_mod5, echo=T}
summary(mdl5)
```

According to the model selected, adding nutrients has a positive effect on fertility (p = 2e-16), whereas clipping plants has a negative effect on fertility (p = 2e-16). Looking at estimates reveals that the positive effect of nutrients is `r round(1.44082/0.45538,2)` times stronger than the negative effect of clipping the plant. The interaction between added nutrients and unclipped is negative (p = 2e-16), meaning that clipping fertilized plants decreases less fertility than clipping plants without added nutrients. This explanation is consistent with what we observe when visualising data __(Figure 3)__. Although the negative effect of clipping on total fruits is weak compared to the positive effect of added nutrients, it is even weaker on fertilized plants. This would mean fertilized plants are more resistant to herbivory.

#Appendices

```{r figure1, fig.cap="Distribution of the response variable, 'total fruits'"}
boxplot(ara$total.fruits)
```

```{r figure2,fig.cap="Fit of total fruit count to different distribution.",echo=F, fig.height=8}

par(mfrow=c(2,2))
qqp(main="Normal",ara$total.fruits,"norm")
poisson <- fitdistr(ara$total.fruits, "Poisson")
qqp(main="Poisson",ara$total.fruits, "pois", poisson$estimate)
qqp(main="Lognormal", ara$total.fruits,"lnorm")

```

```{r figure3, fig.cap="Visualizing interaction between amd and nutrient."}
ggplot(data=ara,aes(x=amd,y=total.fruits))+
  facet_grid(~nutrient,labeller = as_labeller(c(`1`= "No added nutrients",`8`= "Added nutrients")))+
  geom_boxplot()+
  theme_bw()
```

```{r table1}

panderOptions('table.split.table', Inf)
set.caption('Population groups repartition through across regions')
pander(xtabs(~-ara$reg+ara$popu))

#kable(xtabs(~-ara$reg+ara$popu))
```

```{r table2}

panderOptions('table.split.table', Inf)
set.caption('Genotypes repartition across populations')
pander(xtabs(~-ara$reg+ara$popu))

#kable(xtabs(~-ara$reg+ara$popu))
```



