---
title: 'Study of low birth weight'
author: "Cyril Matthey-Doret"
date: "November 3, 2016"
output: pdf_document
linkcolor: "blue"
---

```{r setup, include=F}
library(ggplot2);library(gridExtra)
#setwd("/home/cyril/Documents/Master/sem_1/Data_analysis/birth_weight/")
setwd("/home/cyril/Documents/Master/sem_1/Data_analysis/birth_weight/")
birth_weight<-read.table("birtweigth_study.txt", sep= " ") # data is separated by spaces (even though the format was originally csv)
row.names(birth_weight) <- NULL  # Resetting row names
birth_weight$lwt <- birth_weight$lwt*0.45359237  # Converting from pounds to kg
birth_weight$bwt <- birth_weight$bwt/1000  # Converting from grams to kg
nice_varnames <- c(age = "age [years]", lwt="weight of mother [kg]", bwt="birth weigh [kg]t",ht="known hypertension", ptd="preterm birth", smoke = "smoking mother")

birth_plotter <- function(df){  # Creating a function to plot all variables vs birth weight
  plotlist <- list()  # Initiating the list that will contain plots
  count <- 1  # Counter for list indexing
  for(c in colnames(df)){  # Iterating over variables names
    if(c != "bwt"){
      if(is.numeric(df[1,c])){  # If the variable is numeric, make a scatterplot
        plotlist[[count]] <-ggplot(data=df)+
        geom_point(aes_string(x=df[,c],y=df$bwt))+  # aes_string instead of aes to prevent erasing of values at last iteration
          theme_bw()+xlab(nice_varnames[c])+ylab("Birth weight")+
          ggtitle(paste(LETTERS[count], ": ", nice_varnames[c]))+
          theme(plot.title=element_text(size = 12, face="bold"))
      } else{  # If the variable is a factor, make a boxplot
      plotlist[[count]] <-ggplot(data=df)+
        geom_boxplot(aes_string(x=df[,c],y=df$bwt))+
        theme_bw()+xlab(nice_varnames[c])+ylab("Birth weight")+
        ggtitle(paste(LETTERS[count], ": ", nice_varnames[c]))+
        theme(plot.title=element_text(size = 12, face="bold"))
      }
      count <- count + 1
    }
  }
  return(plotlist)
}

```
#Description of the data:

The dataset contains 6 variables:

  * age: The age of the mother when she gave birth. (Discrete numeric)
  * lwt: The weigth of the mother before pregnancy. (Continuous numeric)
  * smoke: Wether the mother smokes or not. (Two-levels factor)
  * ht: Whether the mother has hypertension or not. (Two-levels factor)
  * bwt: The weight of the newborn. (Continuous numeric)
  * ptd: Wether the newborn is prematured or not. (Two-levels factor)

It is interesting to note that the sample size in each group is strongly unbalanced:

```{r group_size, echo=F}
library(knitr)
kable(summary(birth_weight)[1:2,c(3,4,6)])
```

`r round(100*length(birth_weight$bwt[birth_weight$ht=="HT"])/length(birth_weight$bwt),1)`% of the mothers in the dataset have hypertension, and `r round(100*length(birth_weight$bwt[birth_weight$ptd=="PPREM"])/length(birth_weight$bwt),1)`% of the newborn are prematured. Those values are not representative of real populations. As a matter of fact, in 1989-1994, the prevalence of hypertension in women was estimated to be 23.4% in the USA by the [National Health Examination Surveys](http://emedicine.medscape.com/article/1928048-overview#showall). The percentage of preterm birth in the USA was estimated to 9.6%, in 2014 by the [Center for Disease Control and Prevention](http://www.cdc.gov/nchs/fastats/births.htm).

In addition to poorly representing the general population, this sample does not represent all combination of factors. Although this should be the most represented group, there are no women without hypertension who doesn't smoke and whose child was not born preterm. The absence of such combination will likely affect the reliability of any model when analysing the data. The very different sizes of the groups could also have an impact on the homogeneity of variance.

#Data processing:

Weights units were not provided, but it appears that newborn weights were in pounds and mothers weights were in grams, row numbers also started from 85 and many were missing.
Although we had no formal confirmation of the weight units, we converted them to kilograms.

Visualizing the data reveals that premature babies are heavier than the others (Figure 1, E). These results are surprising and should be interpreted with caution as there may be a mistake in the data. When compared to real populations, both the median values and sample size of the different groups suggest there may have been an inversion of the levels at some point, although it is not possible to be sure of it.

#Statistical analysis:

In order to evaluate the effect of the different factors on the newborn weight, we can try to perform a linear model, since the data is normally distributed (Figure 2).


From the qqplot (Figure 3), the residuals of the model do not seem to deviate too much from the normality. The residuals vs fitted plot does not show a significant pattern or particular signs of heteroscedasticity. 
A linear model should be a reasonable approach. Interactions will not be accounted for in the model as it is complex to infer a biological meaning to interactions between multiple variables.

```{r linear_model, echo=T}

summary(lm(bwt ~ lwt+age+ht+smoke+ptd,data=birth_weight))

```

Since the age is not significant, we can just remove it from the model.

```{r new_model, echo=T}

summary(lm(bwt ~lwt+ht+smoke+ptd,data=birth_weight))

```

Removing the age increased the adjusted R squared. This indicates that the age factor improved the predictory power of our model less than what would be expected by chance.

According to the model, Hypertension status, smoking habit, mother's weight and preterm delivery status will all have an effect on the birth weight, whereas the age of the mother will not.

#Conclusion:

```{r new_modelsum, echo=F}
mdl <- summary(lm(bwt ~lwt+ht+smoke+ptd,data=birth_weight))
```

According to the model, the factors associated with a low birth weight are:

  * Low mother weight (_p-value_ = `r round(as.numeric(mdl$coefficients[,4][2]), 3)`)
  * Non-hypertensive mother (_p-value_ = `r round(as.numeric(mdl$coefficients[,4][3]), 3)`)
  * Full term delivery (_p-value_ = `r round(as.numeric(mdl$coefficients[,4][5]), 3)`)
  * Non-smoking mother (_p-value_ = `r round(as.numeric(mdl$coefficients[,4][4]), 3)`)
  
All of these results are counterintuitive and should be interpreted carefully. The data may have suffered manipulation errors, as mentioned previously, or the surprising results may also be caused by the small sample size in some groups.

#Figures:


```{r dataviz,echo=F, fig.height= 8, fig.width=8, results='asis', fig.cap="Visualization of the different variables against birth weight"}
grid.arrange(grobs=birth_plotter(birth_weight),ncol=2)  # Something is wrong, premature babies are heavier...
```

```{r data_distr, echo=F, fig.height=4, fig.width=6, fig.cap="The birth weight follows a gaussian distribution"}
hist(birth_weight$bwt, breaks=30, main="Distribution of birth weight", xlab="Birth weight [kg]", ylab="Frequency")
```

```{r data_check, echo=F, warning=F, fig.height=8, fig.width=8, fig.cap="Quality check of the linear model"}

par(mfrow=c(2,2))
plot(lm(bwt ~lwt+ht+smoke+ptd,data=birth_weight))

```
