---
title: 'Raw report: Birth weight'
author: "Cyril Matthey-Doret"
date: "November 3, 2016"
output: pdf_document
linkcolor: "blue"
---

```{r setup, include=F}
library(ggplot2);library(gridExtra)
setwd("/home/cyril/Documents/Master/sem_1/Data_analysis/reports/birth_weight/")
birth_weight<-read.table("birtweigth_study.txt", sep= " ") # data is separated by spaces (even though the format was originally csv)
row.names(birth_weight) <- NULL  # Resetting row names
birth_weight$lwt <- birth_weight$lwt*0.45359237  # Converting from pounds to kg
birth_weight$bwt <- birth_weight$bwt/1000  # Converting from grams to kg
nice_varnames <- c(age = "age [years]", lwt="weight of mother [kg]", bwt="birth weigh [kg]t",ht="known hypertension", ptd="preterm birth", smoke = "smoking mother")

birth_plotter <- function(df){  # Creating a function to plot all variables vs birth weight
  plotlist <- list()  # Initiating the list that will contain plots
  count <- 1  # Counter for list indexing
  for(c in colnames(df)){  # Iterating over variables names
    if(c != "bwt"){
      if(is.numeric(df[1,c])){  # If the variable is numeric, make a scatterplot
        plotlist[[count]] <-ggplot(data=df)+
        geom_point(aes_string(x=df[,c],y=df$bwt))+  # aes_string instead of aes to prevent erasing of values at last iteration
          theme_bw()+xlab(nice_varnames[c])+ylab("Birth weight")+
          ggtitle(paste("Newborn weight [kg] VS", nice_varnames[c]))+
          theme(plot.title=element_text(size = 12, face="bold"))
      } else{  # If the variable is a factor, make a boxplot
      plotlist[[count]] <-ggplot(data=df)+
        geom_boxplot(aes_string(x=df[,c],y=df$bwt))+
        theme_bw()+xlab(nice_varnames[c])+ylab("Birth weight")+
        ggtitle(paste("Newborn weight [kg] VS", nice_varnames[c]))+
        theme(plot.title=element_text(size = 12, face="bold"))
      }
      count <- count + 1
    }
  }
  return(plotlist)
}

```
#Description of the data:

The dataset contains 6 variables:

  * age: The age of the mother when she gave birth. (Discrete numeric)
  * lwt: The weigth of the mother before pregnancy. (Continuous numeric)
  * smoke: Wether the mother smokes or not. (Two-levels factor)
  * ht: Whether the mother has hypertension or not. (Two-levels factor)
  * bwt: The weight of the newborn. (Continuous numeric)
  * ptd: Wether the newborn is prematured or not. (Two-levels factor)

It is interesting to note that the sample size is very small for some factor levels:


#Data processing:

Weights units were not provided, but it appears that newborn weights were in pounds and mothers weights were in grams, row numbers also started from 85 and many were missing.
Although we had no formal confirmation of the weight units, we converted them to kilograms.

Visualizing the data reveals that premature babies are heavier than the others (Figure 1). These results are surprising and should be interpreted with caution as there may be a mistake in the data. The factor levels may for example have been switched.

There is a weak correlation between the weight of mother and the weight of the child (p-value = `r round(cor.test(birth_weight$lwt, birth_weight$bwt)$p.value, 3)`, coefficient = `r round( as.numeric(cor(birth_weight$bwt, birth_weight$lwt)), 3)`)

```{r dataviz,echo=F, fig.height= 8, fig.width=8}
grid.arrange(grobs=birth_plotter(birth_weight),ncol=2)  # Something is wrong, premature babies are heavier...
```

In order to measure the impact of the different factors on the newborn weight, we can try to perform a linear model.

```{r data_check, echo=F, warning=F, fig.height=8}


#qqnorm(resid(lm(bwt ~ ht*smoke*lwt*ptd*age,birth_weight)))
#qqline(resid(lm(bwt ~ ht*smoke*lwt*ptd*age,birth_weight)))
par(mfrow=c(2,2))
plot(lm(bwt ~ ht*smoke*lwt*ptd*age,birth_weight))

#shapiro.test(residuals(lm(bwt ~ ht*smoke*lwt*ptd*age,birth_weight)))


```

From the qqplot, the residuals of the model do not seem to deviate too much from the normality. The residuals vs fitted plot does not show a significant pattern.
A linear model should provide a reasonable approximation.

```{r linear_model, echo=T}
summary(lm(bwt ~ ht+smoke+lwt+ptd+age,birth_weight))
summary(lm(bwt ~ ht+smoke+lwt+ptd,birth_weight))
```


