---
title: 'Raw report: Birth weight'
author: "Cyril Matthey-Doret"
date: "November 3, 2016"
output: pdf_document
linkcolor: "blue"
---

```{r setup, include=F}
library(ggplot2);library(gridExtra)
birth_weight<-read.table("birtweigth_study.txt", sep= " ") # data is separated by spaces (even though the format was originally csv)
row.names(birth_weight) <- NULL  # Resetting row names
birth_weight$lwt <- birth_weight$lwt*0.45359237  # Converting from pounds to kg
birth_weight$bwt <- birth_weight$bwt/1000  # Converting from grams to kg
nice_varnames <- c(age = "age [years]", lwt="weight of mother [kg]", bwt="birth weigh [kg]t",ht="known hypertension", ptd="preterm birth", smoke = "smoking mother")

birth_plotter <- function(df){  # Creating a function to plot all variables vs birth weight
  plotlist <- list()  # Initiating the list that will contain plots
  count <- 1  # Counter for list indexing
  for(c in colnames(df)){  # Iterating over variables names
    if(c != "bwt"){
      if(is.numeric(df[1,c])){  # If the variable is numeric, make a scatterplot
        plotlist[[count]] <-ggplot(data=df)+
        geom_point(aes_string(x=df[,c],y=df$bwt))+  # aes_string instead of aes to prevent erasing of values at last iteration
          theme_bw()+xlab(nice_varnames[c])+ylab("Birth weight")+
          ggtitle(paste("Newborn weight [kg] VS", nice_varnames[c]))+
          theme(plot.title=element_text(size = 12, face="bold"))
      } else{  # If the variable is a factor, make a boxplot
      plotlist[[count]] <-ggplot(data=df)+
        geom_boxplot(aes_string(x=df[,c],y=df$bwt))+
        theme_bw()+xlab(nice_varnames[c])+ylab("Birth weight")+
        ggtitle(paste("Newborn weight [kg] VS", nice_varnames[c]))+
        theme(plot.title=element_text(size = 12, face="bold"))
      }
      count <- count + 1
    }
  }
  return(plotlist)
}

```
#Data processing:

Weights were in different units (kg and pounds), row numbers started from 85 and many were missing.

Visualising the data reveals that premature babies are heavier than the others (Figure 1). These results are surprising and should be interpreted with caution as there may be a mistake in the data. The factor levels may for example have been switched.
If we consider the data to be valid anyway and analyze it, we can see the residuals are not normally distributed, but a square transformation of the birth weight values solved this issue.
After transforming the data, we performed an ANCOVA.

```{r dataviz,echo=F, fig.height= 8, fig.width=8}
grid.arrange(grobs=birth_plotter(birth_weight),ncol=2)  # Something is wrong, premature babies are heavier...
```


```{r data_check, echo=F}

qqnorm(birth_weight$bwt)
qqline(birth_weight$bwt)
shapiro.test(residuals(lm(bwt ~ ht*smoke*lwt*ptd*age,birth_weight)))
shapiro.test(residuals(lm(bwt^2 ~ ht*smoke*lwt*ptd*age,birth_weight)))

```




