---
title: 'Raw report: Birth weight'
author: "Cyril Matthey-Doret"
date: "November 3, 2016"
output: pdf_document
---

```{r setup, include=F}
library(ggplot2);library(gridExtra)
birth_weight<-read.table("birtweigth_study.txt", sep= " ") # data is separated by spaces (even though the format was originally csv)
row.names(birth_weight) <- NULL  # Resetting row names
birth_weight$lwt <- birth_weight$lwt*0.45359237  # Converting from pounds to kg
birth_weight$bwt <- birth_weight$bwt/1000  # Converting from grams to kg
nice_varnames <- c(age = "age [years]", lwt="weight of mother [kg]", bwt="birth weigh [kg]t",ht="known hypertension", ptd="preterm birth", smoke = "smoking mother")
birth_plotter <- function(df){  # Creating a function to plot all variables vs birth weight
  plotlist <- list()  # Initiating the list that will contain plots
  count <- 1  # Counter for list indexing
  for(c in colnames(df)){  # Iterating over variables names
    if(c != "bwt"){
      if(is.numeric(df[1,c])){  # If the variable is numeric, make a scatterplot
        plotlist[[count]] <-ggplot(data=df)+
        geom_point(aes_string(x=df[,c],y=df$bwt))+  # aes_string instead of aes to prevent erasing of values at last iteration
          theme_bw()+xlab(nice_varnames[c])+ylab("Birth weight")
      } else{  # If the variable is a factor, make a boxplot
      plotlist[[count]] <-ggplot(data=df)+
        geom_boxplot(aes_string(x=df[,c],y=df$bwt))+
        theme_bw()+xlab(nice_varnames[c])+ylab("Birth weight")
      }
      count <- count + 1
    }
  }
  return(plotlist)
}

```

Visualising the date reveals that premature babies are heavier than the others (Figure 1). These results are surprising and should be interpreted with caution as there may be a mistake in the data. The factor levels may for example have been switched.
If we consider the data to be valid anyway and analyze it, we can build a linear model and perform an ANOVA on it, as the data is normally distributed (Figure 2).

```{r dataviz,echo=F, fig.height= 8}
grid.arrange(grobs=birth_plotter(birth_weight),ncol=2)  # Something is wrong, premature babies are heavier...
```


```{r data_check, echo=F}

qqnorm(birth_weight$bwt)
qqline(birth_weight$bwt)

```



#### Comment de Virginie : Pour ton dernier paragraphe, c'est pas une ANCOVA qu'on devrait faire à la place d'une ANOVA? Vu qu'on a plusieurs variables qui ont le potentiel d'influencer la variable réponse ?
